---
title: "Weather and Corn Yield Regressions"
author: "Orion Cardenas and Nathan Mueller"
date: "2/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
library(dplyr)
library(gridExtra)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```
```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r winter temps}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")

```

### Download NASS corn yield data

```{r yield download, include=FALSE}

# set our API key with NASS
nassqs_auth(key = "4D18D90E-0607-33B1-9600-269C28AC265A")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

There is a significant trend in yield over time. Yield has increased over time and the low p values and positive intercept indicates that it is a significant positive trend.

```{r}
winne_corn_yields = cornyields %>%
  filter(county_name == "WINNESHIEK") %>%
  group_by(year)

ggplot(winne_corn_yields, mapping = aes(x = year, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Year", y = "Corn Yield (Bu/Acre)") +
  ggtitle("Yield Over Time") +
  geom_smooth(method = lm)

lm_winne_corn_yields = lm(yield ~ year, winne_corn_yields)

summary(lm_winne_corn_yields)
```


### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

No, there is no evidence for slowing yield growth. After fitting a quadratic time trend, there appears to be a positive quadratic trend indicating increasing growth over time.

```{r}
winne_corn_yields$yearsq = winne_corn_yields$year^2

lm_wcy_quad = lm(yield ~ year + yearsq, winne_corn_yields)
summary(lm_wcy_quad)
winne_corn_yields$fitted = lm_wcy_quad$fitted.values

ggplot(winne_corn_yields) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  ggtitle("Quadratic Time Trend") +
  labs(x = "Year", y = "Corn Yield (Bu/Acre)")

```


### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

Summer avg Tmax alone does not appear to be a good predictor of corn yield (p = 0.29). Adding both year and summer avg Tmax does improve the model. The model with year and avg T max had the most significant p value although the meantmax variable was not significant. The quadratic model was significant (p=0.007) and all of the variables in the model were good predictors of meantmax.

```{r}
winne_yield_summertmax = inner_join(winne_corn_yields, winnesummer)

winne_yield_summertmax$meantmaxq = winne_yield_summertmax$meantmax^2

lm_winne_yieldtmax = lm(yield ~ meantmax, winne_yield_summertmax)

lm_winne_yieldtmaxyear = lm(yield ~ meantmax + year, winne_yield_summertmax)

lm_winne_yieldtmaxq = lm(yield ~ meantmax + meantmaxq, winne_yield_summertmax)

summary(lm_winne_yieldtmax)
summary(lm_winne_yieldtmaxyear)
summary(lm_winne_yieldtmaxq)

ggplot(winne_yield_summertmax, mapping = aes(x = meantmax, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Mean T Max", y = "Corn Yield (BU/Acre)") +
  ggtitle("Yield by MeanTMax") +
  geom_smooth(method = lm)
```


### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

Meantmax summer is a fair predictor of corn yield across all counties (p=0.06). The model improves when adding the quadratic variable (p=0.001). Meantmax winter was not a good model (p= 0.9). The performance of the meantmax winter model was drastically improved when the quadratic variable was added (p=0.0005). The combination of meantmax summer and meantmax winter produced a good model where both variables were good predictors (p=0.0004). It was also greatly improved with the addition of the quadratic variables.

```{r}
allcounty_corn_yields_2018 = cornyields[cornyields$year == 2018,]
allcounty_corn_yields_2018$countyfp = as.factor(allcounty_corn_yields_2018$county_ansi)

allcounty_summer_2018 = tmaxdf %>%
  filter(year == 2018 & doy >= 152 & doy <= 243) %>%
  group_by(year, countyfp) %>%
  summarize(meantmaxsum = mean(tmax))

allcounty_winter_2018 = tmaxdf %>%
  filter((doy <= 59) & !is.na(tmax)) %>%
  filter(year == 2018) %>%
  group_by(year, countyfp) %>%
  summarize(meantmaxwin = mean(tmax))

allcounty_corn_yields_2018 = left_join(allcounty_corn_yields_2018, allcounty_summer_2018, keep=F, by = "countyfp")

allcounty_corn_yields_2018 = left_join(allcounty_corn_yields_2018, allcounty_winter_2018, keep=F, by = "countyfp")

allcounty_corn_yields_2018$meantmaxsumq = allcounty_corn_yields_2018$meantmaxsum^2

allcounty_corn_yields_2018$meantmaxwinq = allcounty_corn_yields_2018$meantmaxwin^2

lm_allcy_tempyield_sum = lm(yield ~ meantmaxsum, allcounty_corn_yields_2018)

lm_allcy_tempyield_sumq = lm(yield ~ meantmaxsum + meantmaxsumq, allcounty_corn_yields_2018)


lm_allcy_tempyield_win = lm(yield ~ meantmaxwin, allcounty_corn_yields_2018)

lm_allcy_tempyield_winq = lm(yield ~ meantmaxwin + meantmaxwinq, allcounty_corn_yields_2018)

lm_allcy_tempyield_sumwin1 = lm(yield ~ meantmaxsum + meantmaxwin, allcounty_corn_yields_2018)

lm_allcy_tempyield_sumwin2 = lm(yield ~ meantmaxsum + meantmaxwin + meantmaxsumq + meantmaxwinq, allcounty_corn_yields_2018)

summary(lm_allcy_tempyield_sum)
summary(lm_allcy_tempyield_sumq)
summary(lm_allcy_tempyield_win)
summary(lm_allcy_tempyield_winq)
summary(lm_allcy_tempyield_sumwin1)
summary(lm_allcy_tempyield_sumwin2)




AIC(lm_allcy_tempyield_sum, lm_allcy_tempyield_sumq, lm_allcy_tempyield_win, lm_allcy_tempyield_winq, lm_allcy_tempyield_sumwin1, lm_allcy_tempyield_sumwin2)

sumyieldplot2018 = ggplot(allcounty_corn_yields_2018, mapping = aes(x = meantmaxsum, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Summer Mean Daily Maximum Temperature", y = "Yield") +
  ggtitle("Yield by Summer MeanTmax for All Counties in 2018") +
  geom_smooth(method = lm)

winyieldplot2018 = ggplot(allcounty_corn_yields_2018, mapping = aes(x = meantmaxwin, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Winter Mean Daily Maximum Temperature", y = "Yield") +
  ggtitle("Yield by Winter MeanTmax for All Counties in 2018") +
  geom_smooth(method = lm)

grid.arrange(sumyieldplot2018,winyieldplot2018)
```


### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

Including the county variable in our regression results in a strong model. The two temperature variables (Tmax and Tmax^2) are highly significant in this model (p<2.2e-16). They are better predictors of corn yield than the county variables.

```{r}
cornyields$countyfp = as.factor(cornyields$county_ansi)

allcounty_summer_ay = tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(year, countyfp) %>%
  summarize(meantmax = mean(tmax))

allcounty_cy_summertemp = inner_join(cornyields, allcounty_summer_ay, by=c("countyfp", "year")) %>%
  na.omit(.)

allcounty_cy_summertemp$meantmaxq = allcounty_cy_summertemp$meantmax^2

lm_allcy_summertempyield = lm(yield ~ countyfp + meantmax + meantmaxq, allcounty_cy_summertemp)
summary(lm_allcy_summertempyield)

allcounty_cy_summertemp$fitted = lm_allcy_summertempyield$fitted.values

actualvfittedplot1 = ggplot(allcounty_cy_summertemp) +
      geom_point(mapping = aes(x = meantmax, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = meantmax, y = fitted, col="Fitted")) +
      theme_bw() +
      labs(x = "Average Daily Maximum Temperature (°C)", y = "Corn Yield (Bu/Acre)") +
      ggtitle("Actual Vs Fitted Corn Values By Temperature (°C)")

actualvfittedplot2 = ggplot(allcounty_cy_summertemp) +
      geom_point(mapping = aes(x = year, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = year, y = fitted, col="Fitted")) +
      theme_bw() +
      labs(x = "Year", y = "Corn Yield (Bu/Acre)") +
      ggtitle("Actual Vs Fitted Corn Values Over Time")
grid.arrange(actualvfittedplot1, actualvfittedplot2)
```

### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

I decided to do a panel for soybean yield across all Iowa counties and years. The model is strong overall (p < 2e-16) and the temperature variables are highly significant in this model (p < 2e-16). 

```{r, include = FALSE}

# set our API key with NASS
nassqs_auth(key = "4D18D90E-0607-33B1-9600-269C28AC265A")

# parameters to query on 
params = list(commodity_desc = "SOYBEANS", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
soybeanyieldsall = nassqs_yields(params)

soybeanyieldsall$county_ansi = as.numeric(soybeanyieldsall$county_ansi)
soybeanyieldsall$yield = as.numeric(soybeanyieldsall$Value)

# clean and filter this dataset
soybeanyields = select(soybeanyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soybeanyields =  tibble(soybeanyields)
```


```{r}
# panel study
soybeanyields$countyfp = as.factor(soybeanyields$county_ansi)

allcounty_summer_ay = tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(year, countyfp) %>%
  summarize(meantmax = mean(tmax))

allcounty_sy_summertemp = inner_join(soybeanyields, allcounty_summer_ay, by=c("countyfp", "year")) %>%
  na.omit(.)

allcounty_sy_summertemp$meantmaxq = allcounty_sy_summertemp$meantmax^2

lm_allsy_summertempyield = lm(yield ~ countyfp + meantmax + meantmaxq, allcounty_sy_summertemp)
summary(lm_allsy_summertempyield)

allcounty_sy_summertemp$fitted = lm_allsy_summertempyield$fitted.values

actualvfittedplot3 = ggplot(allcounty_sy_summertemp) +
      geom_point(mapping = aes(x = meantmax, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = meantmax, y = fitted, col="Fitted")) +
      theme_bw() +
      labs(x = "Average Daily Maximum Temperature (°C)", y = "Soybean Yield") +
      ggtitle("Actual Vs Fitted Soybean Values By Temperature (°C)")

actualvfittedplot4 = ggplot(allcounty_sy_summertemp) +
      geom_point(mapping = aes(x = year, y = yield, col="Actual")) +
      geom_line(mapping = aes(x = year, y = fitted, col="Fitted")) +
      theme_bw() +
      labs(x = "Year", y = "Soybean Yield") +
      ggtitle("Actual Vs Fitted Soybean Values Over Time")
grid.arrange(actualvfittedplot3, actualvfittedplot4)

```

